\thispagestyle{plain}
\begin{center}{\huge\textit{Acknowledgements}\par}\end{center}
\vspace{0.5cm}
\addchaptertocentry{\acknowledgementname} % Add the Acknowledgements to the table of contents

\label{chap:acknowledgement}
First of all, I would like to express my sincere gratitude to my thesis supervisor, Prof. Jean Vanderdonckt, for providing me with the opportunity to complete this thesis. He has always supported me throughout my Ph.D. and related research. His patience, motivation, and extensive knowledge have been incredibly valuable. In addition, his advice helped me throughout the research and writing of this thesis.\\

I would like to express my gratitude to my co-supervisor, Professor Paolo Roselli, for his invaluable collaboration and assistance in navigating the complex knowledge of geometric algebra and its application in gesture recognition. His guidance and encouragement have played a pivotal role in shaping the direction of my research.\\

I would like to express my gratitude to the members of my Ph.D. committee for reviewing the manuscript and providing valuable comments. To Prof. Rachid Gherbi, whom I met during my master's degree in Algeria and who was a source of inspiration for his dedication to transmitting his knowledge about the VR/AR field by giving constructive advice. I would like to thank Prof. Suzanne Kieffer and Prof. Giovanni Saggio for their expertise and contribution to the quality of this thesis. I would also like to thank Prof. Charles Pecheur for being the president of my thesis committee.\\

Special thanks to Prof. Radu Vatavu, from Stefan cel Mare University of Suceava, who shared his deep knowledge in gesture interaction~\cite{Vatavu:2023:HHCI} with me at different stages of my research.

In addition, I would like to thank my colleagues at the Louvain Research Institute in Management and Organizations (LouRIM) Institute and the assistants and researchers for their fruitful discussions, comments, and support throughout this research and teaching work. I would also like to thank the administrative team for their invaluable advice and excellent work in facilitating our tasks and creating a favorable working environment.\\

Finally, I am deeply grateful to my family and friends for their unconditional love, understanding, and support throughout this Ph.D. journey. I want to express my profound gratitude for their constant source of motivation, strength, and invaluable advice.



